\documentclass[twocolumn,superscriptaddress,prb]{revtex4-2}
% Alternative: \documentclass[preprint,aps,prb]{revtex4-2}

\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{siunitx}
\graphicspath{{figures/}}

\newcommand{\dart}{\textsc{Dart}}
\newcommand{\ef}{E_\mathrm{f}}
\newcommand{\mae}{\mathrm{MAE}}
\newcommand{\eV}{\,\mathrm{eV}}

\begin{document}

\title{DART: A Compact Defect-Aware Radial Transformer for High-Throughput Prediction of Defect Formation Energies in Two-Dimensional Materials}

\author{Yiming Hua}
\affiliation{Wuhan University, Wuhan, China}
% TODO: add other authors

\date{\today}

\begin{abstract}
Defect engineering in two-dimensional (2D) materials demands rapid screening
of defect formation energies (\ef) across vast combinatorial spaces, yet
density functional theory (DFT) calculations remain prohibitively expensive.
Existing graph neural network (GNN) surrogates achieve reasonable accuracy on
random test splits but lack systematic evaluation of generalization to unseen
host materials---a prerequisite for trustworthy deployment.
%
Here we introduce \dart{} (Defect-Aware Radial Transformer), a compact
0.75\,M-parameter architecture coupling local bond--angle message passing
with global periodic self-attention biased by minimum-image distances, and
augmented by an explicit defect-site embedding.
%
On the IMP2D benchmark (10\,641 DFT-converged impurity configurations in 44
host monolayers), \dart{} achieves a single-model mean absolute error (MAE)
of 0.407\eV{} and an ensemble MAE of 0.359\eV{}, reducing the error by 25\%
and 34\% relative to ALIGNN (4.03\,M parameters), respectively.
%
We further propose a three-tier \emph{Graduated Generalization Assessment}:
(i)~in-distribution random split, (ii)~intra-family out-of-distribution via
leave-one-host-out cross-validation over G6 transition-metal dichalcogenides,
and (iii)~compositional out-of-distribution via block-out matrix completion.
%
\dart{} degrades gracefully from 0.36 to 0.54\eV{} (1.5$\times$) under
intra-family extrapolation---far milder than the 4.9$\times$ naive-baseline
gap---demonstrating robust transferability within the TMD family.
%
Temperature-scaled ensemble uncertainty achieves 93.4\% empirical coverage at
the 90\% nominal level, enabling reliable high-throughput screening with
calibrated confidence bounds.
\end{abstract}

\maketitle

%----------------------------------------------------------------------
\section{Introduction}\label{sec:intro}
%----------------------------------------------------------------------
\input{sections/introduction}

%----------------------------------------------------------------------
\section{Method}\label{sec:method}
%----------------------------------------------------------------------
\input{sections/method}

%----------------------------------------------------------------------
\section{Experimental Setup}\label{sec:setup}
%----------------------------------------------------------------------
\input{sections/setup}

%----------------------------------------------------------------------
\section{Results}\label{sec:results}
%----------------------------------------------------------------------
\input{sections/results}

%----------------------------------------------------------------------
\section{Discussion}\label{sec:discussion}
%----------------------------------------------------------------------
\input{sections/discussion}

%----------------------------------------------------------------------
\section{Conclusion}\label{sec:conclusion}
%----------------------------------------------------------------------
\input{sections/conclusion}

\begin{acknowledgments}
This work was supported by the National College Students Innovation and
Entrepreneurship Training Program (Key Support Area Project).
Computations were performed on 8$\times$NVIDIA L40S GPUs at Wuhan University.
\end{acknowledgments}

\bibliography{references}

\end{document}
