% !TEX root = ../main.tex

%----------------------------------------------------------------------
\subsection{Why a compact model suffices}
%----------------------------------------------------------------------

\dart{}'s 0.75\,M parameters outperform ALIGNN's 4.03\,M by a substantial
margin.
This result aligns with emerging evidence that, in the data-limited regime
characteristic of materials science ($N \sim 10^3$--$10^4$), training
strategy and data engineering dominate over architectural complexity.
Our scaling analysis (not shown) finds a power-law exponent
$\alpha = -0.40$ for data scaling but $\beta \approx -0.01$ for parameter
scaling ($R^2 = 0.95$), confirming that \emph{more data, not more
parameters}, is the primary lever for improvement on this task.

\paragraph{Decomposing the improvement.}
The comparison between \dart{} (0.407\eV{}) and the vanilla
CrystalTransformer backbone (0.516\eV{}) trained with the same
architecture but a standard MSE/plateau recipe directly quantifies the
contribution of our training and data engineering.
Of the total 0.133\eV{} improvement, the training pipeline
(MAE loss, cosine warmup, SWA) accounts for ${\sim}0.10$\eV{}
(Table~\ref{tab:ablation}), while the defect-site embedding and
ct-UAE transfer embeddings contribute the remaining
${\sim}0.03$\eV{}.
This breakdown reinforces our thesis: in the small-data regime,
\emph{how} one trains matters at least as much as \emph{what}
one trains.

%----------------------------------------------------------------------
\subsection{Self-attention as learned potential summation}
%----------------------------------------------------------------------

The success of the global geometric attention layers can be understood
through the lens of ``neural potential summation''~\cite{crystalformer2024}:
the learned distance bias $\phi_\mathrm{bias}(d_{ij}^{\mathrm{PBC}})$ acts
as a non-parametric interatomic potential that gates information flow.
In solid-state physics, the crystal potential is naturally decomposed into
Fourier components in reciprocal space; analogously, the Gaussian RBF
expansion that feeds $\phi_\mathrm{bias}$ provides a localized basis in
real space from which the MLP learns smooth, physically meaningful coupling
functions.
For defect systems, where the perturbation source (dopant) induces
long-range elastic strain fields decaying as ${\sim}1/r^2$ in 2D, this
mechanism naturally allocates higher attention weights to nearby atoms while
retaining non-zero coupling to distant atoms---precisely the physics of
defect-induced perturbation.

The attention-weight analysis (\S\ref{sec:interpret}) provides empirical
confirmation: the defect atom emerges as a global attention hub without
explicit architectural enforcement, consistent with its role as the dominant
perturbation source.

%----------------------------------------------------------------------
\subsection{Methodological value of graduated OOD assessment}
%----------------------------------------------------------------------

Our three-tier protocol addresses a blind spot identified by
Li~\textit{et al.}~\cite{li2025ood}: most materials ML papers report only
random-split performance, which tests interpolation rather than genuine
extrapolation.
By systematically varying the degree of distributional shift---from
in-distribution to intra-family host transfer to compositional
extrapolation---we provide a nuanced picture of model reliability.

We recommend that future materials ML studies adopt similar tiered
protocols, with at least one OOD tier beyond random splitting.
The LOHO framework is readily transferable to other multi-component
material families (e.g., high-entropy alloys, hybrid perovskites).

%----------------------------------------------------------------------
\subsection{Toward practical defect engineering}
%----------------------------------------------------------------------

Beyond benchmarking accuracy, \dart{}'s calibrated predictions enable
concrete materials-design workflows.
Given that low-$\ef$ dopants are thermodynamically accessible under
typical synthesis conditions, a practitioner can use \dart{} to rapidly
rank all 65 candidate dopants for a new TMD host and identify the
top-$k$ candidates for experimental validation---reducing the DFT
pre-screening burden from thousands of CPU-hours to seconds of GPU
inference.
%
Moreover, the per-dopant error map (Fig.~\ref{fig:periodic_table})
reveals that predictions for main-group dopants (Bi, Pb, Sr) are
already reliable enough (MAE $< 0.17$\eV{}) for semi-quantitative
screening, while early transition-metal dopants (Sc, Mn, Ta) require
additional caution.
This element-resolved confidence map can directly guide experimentalists:
high-confidence predictions warrant immediate synthesis attempts, while
uncertain predictions should be verified by targeted DFT calculations
or prioritized in active-learning loops.
%
In the context of defect engineering for device applications---where
beneficial defects (e.g., luminescent centers for single-photon emitters,
shallow dopants for carrier modulation) must be distinguished from
detrimental ones (deep traps, recombination centers)---\dart{}'s
ability to rapidly evaluate $\ef$ across the full host--dopant matrix
provides a first-pass filter before more expensive excited-state or
transport calculations.

%----------------------------------------------------------------------
\subsection{Limitations and outlook}
%----------------------------------------------------------------------

Several limitations warrant acknowledgment:
(1)~The current model covers only the 44 host materials in IMP2D;
generalization to entirely new material families (e.g., MXenes, h-BN
alloys) requires expanded training data.
(2)~The IMP2D database contains only substitutional adsorbates and
interstitials; vacancy-type defects---single vacancies, divacancies, and
Stone--Wales reconstructions---are absent from both training and
evaluation.
Incorporating vacancy datasets, such as existing graphene defect
libraries containing hundreds of defect configurations, would extend
\dart{}'s applicability to the broader defect landscape.
(3)~We predict only defect formation energies; extending to other
defect-related properties---charge transition levels, migration barriers,
carrier capture coefficients, and optical transition energies---would
broaden the platform's utility for full defect engineering pipelines.
(4)~The OOD evaluation is limited to the G6-TMD subfamily; a broader
cross-family assessment awaits datasets with sufficient diversity.

Looking forward, the integration of \dart{} with active learning loops
could efficiently expand the training distribution into unexplored
regions of host--dopant space, guided by the calibrated uncertainty
estimates demonstrated here.
Coupling \dart{} predictions with high-throughput transport and optical
simulations would enable end-to-end screening of dopants for targeted
device functionalities.
